PACE 2024: OCMu64, reductions and a branch-and-bound solver for\\ One-sided Crossing Minimization

### 2024-09-01 tex/poster.tex
simplify

\title{OCMu64: a solver for One-sided Crossing Minimization}

@@

@@

  \begin{alertblock}{todo}
    increase font size? Make lemmas inside blocks, not lemmas as blocks. Figures.
  \end{alertblock}

  \begin{block}{Problem statement}
The 2024 edition of PACE, an annual optimization challenge, considers the
\emph{one-sided crossing minimization} problem, defined as follows.
Given is a bipartite graph $(A, B)$ that is drawn in the plane at points
$(i, 1)$ and $(j,0)$ for components $A$ and $B$ respectively. The ordering of $A$
is fixed, and the goal is the find an ordering of $B$ that minimizes the number of
crossings when edges are drawn as straight lines.

We introduce some new reductions and give an overview of our algorithm. Proofs are brief or
omitted due to lack of space.

\heading{Definitions}
We use $<$ to compare vertices in $A$ in their fixed ordering. We generalize to weighted graphs
for the proof of \cref{lem:sfp}: a node $u \in B$ is taken to be a function $u: A \to \mathbb
R^{\geq 0}$, where weight 0 represents an absent edge. We write $W_u = \sum_{a \in A}u(a)$ for
the total weight of $u$, and set $\bar u = u/W_u$. Write $N(u) = \supp(u) \subseteq A$ for the
set of neighbors of $u$.

We write $c(u, v) = \sum_{(a, b) \in A^2}
u(a)v(b)[a > b]$; in the unweighted case, this is the \emph{crossing number}, the number of
crossings between edges incident to $u$ and $v$ when $u$ is drawn before $v$.  For
$X,Y\subseteq B$ we set $c(X,Y) = \sum_{x\in X}\sum_{y\in Y} c(x,y)$ for the cost of ordering
all vertices of $X$ before all vertices of $Y$. More generally, $c(X,Y,Z) = c(X, Y) + c(X, Z) +
c(Y, Z)$. We also consider the \emph{reduced cost} $r(X,Y) = c(X, Y) - c(Y, X)$, which is
negative when $X$ is better \emph{before} $Y$ and positive when $X$ is better \emph{after} $Y$.

We write $u\b v$ when $u$ must come before $v$ in all minimal solutions, and say that $(u, v)$
is a \emph{fixed} pair. The following result is well known \cite[Lemma
  1]{dujmovic_2004}, RR1 of \cite{dujmovic_2008}.
  \end{block}

\begin{block}{Strongly fixed pairs}
    We call $u, v \in B$ a \emph{strongly fixed pair} if for every $b \in A$ we have ${\sum_{a
    \leq b} \bar u(a) \geq \sum_{a \leq b} \bar v(a)}$, and at least one of these inequalities
    is strict. Note that this implies $r(u, v) < 0$.

    Consider $u \neq v$ from the original, unweighted problem, taking only values $0, 1$. Let
    $n = |N(u)|, m = |N(v)|$, and consider both as ordered lists. Then $u, v$ are strongly
    fixed if and only if for all $0 \leq i < n$, $N(u)_i \leq N(v)_{\lfloor i\cdot m/n
    \rfloor}$ and at least one of the inequalities is strict, which is how we check in practice whether $u, v$ is strongly fixed.

    \textbf{Main lemma:}
    If $(u, v)$ is strongly fixed, then $u \b v$.

    \textbf{The lamma is optimal:}
    Suppose that $u, v \in B$ with $\bar u \neq \bar v$ are not strongly fixed, and let $b \in
    A$ be some element such
    that $\sum_{a \leq b} \bar u(a) > \sum_{a \leq b} \bar v(a)$. By taking $X: A \to \mathbb
    R^{\geq 0}$ some function which assigns very high
    weight to $b$, we can obtain $c(v, X, u) < \min(c(X, u, v), c(u, v, X))$.
\end{block}

  \begin{block}{Practically fixed pairs}
Although such a function $X$ may exist in theory, it does
not have to exist in the actual set $B$, motivating the following definition
that generalizes RRLO2 of \cite{dujmovic_2008}.

\textbf{Blocking set:}
  Suppose $r(u,v)< 0$.
    A \emph{blocking set} $X\subseteq B-\{u,v\}$ is a set such that $c(v,X,u) \leq \min(c(v, u,
    X), (X, v, u))$.  If there is no blocking set for $(u, v)$, we call it a
    \emph{practically fixed pair}, and $u\b v$.
  \end{block}

@@

Note that we do not require $(v, X, u)$ to be a true local minimum, since we do
not consider interactions between vertices in $X$, as that would make ruling out the
existence of such sets much harder.

@@

\begin{column}{\colwidth}
\begin{block}{Gluing}
We now turn our attention to \emph{gluing}, i.e., proving that two vertices $u$
and $v$ always go right next to each other, and we can treat them as a single vertex. First,
let us see that we cannot get `strong gluing' result analogous to \cref{lem:sfimplf}.

\textbf{Remark:}
  When $N(u)=N(v)$ in the unweighted case, or more generally $\bar u=\bar v$, we can glue $u$
    and $v$.
  Otherwise when $r(u,v)\leq 0$, there is an $X : A\to \mathbb R^{\geq 0}$ such that $(u, X, v)$ is
  strictly better than $(u,v,X)$ and $(X,u,v)$.

\textbf{Practical gluing lemma:}
    Let $u$ and $v$ satisfy $r(u, v) \leq 0$.
    A non-empty subset $X\subseteq B-\{u,v\}$ is \emph{blocking} when $c(u, X, v) \leq
    \min(c(u, v, X), c(X, u, v))$. If there is no blocking set, then we can glue $u, v$.

Again such sets $X$ can be found or proven to not exist using a knapsack
algorithm: add points $P_x = (r(u, x), r(x, v))$ and search for a non-empty
set summing to $\leq{}(0,0)$.

\textbf{Greedy gluing}
  When $r(u, x)\leq 0$ for all $x\in B$, there is a solution that
  starts with $u$.
\end{block}

  \begin{block}{Solver}

Our solver \texttt{OCMu64} is based on a standard branch-and-bound on the ordering of the
solution.  We start with fixed prefix $P=()$ and tail $T=B$, and in each step we try (a
subset of) all vertices in $T$ as the next vertex appended to $P$.
In a preprocessing step we compute the trivial lower bound $S_0 =
\sum_{u,v}\min(c(u,v),c(v,u))$ on the score.
We keep track of the score $S_P$ of the prefix and $S_{PT}=c(P, T)$ of
prefix-tail intersections, and abort when this score goes above the best
solution found so far. The \emph{excess} of a tail is its optimal score minus
the trivial lower bound. We do a number of optimizations.

\begin{description}
  \item[Graph simplification] We drop degree-$0$ vertices, merge identical
    vertices, and split the graph into \emph{independent} components
    \cite[Corollary 2]{dujmovic_2004} when
    possible. We find an initial solution using the median heuristic \cite{eades_median,eades_median_2} and a local search that tries to move
    slices and optimally insert them \cite{makinen_experiments,eades_heuristics}, and re-label all nodes accordingly to make
    memory accesses more efficient.
  \item[Fixed pairs] We find all strongly fixed pairs and store them. For the
    exact track we also find practically fixed pairs. Instances for the parameterized
    track are simple enough that the overhead was not worth it. Also for each
    tail we search for new `tail-local' practically fixed pairs. In each state, we only
    try vertices $u\in T$ not fixed by another $v\in T$.
  \item[Gluing] We use the greedy strategy of \cref{greedy}. Our implementation
    of \cref{pg} contained a bug, so we did not use this. (Also benefits
    seemed limited.)
  \item[Tail cache] In each step, we search for the longest suffix of $T$ that
    was seen before, and reuse (the lower bound on) its excess. We also
    cache the tail-local practically fixed pairs.
  \item[Optimal insert] Instead of simply appending $u$ to $P$, we
    insert it in the optimal position. Note that the
    implementation is tricky because it interacts in complicated ways with
    the caching of results for each tail.
\end{description}
  \end{block}

  \begin{block}{Performance}
    \
  \end{block}

>>>>

\title{PACE 2024: OCMu64, reductions and a branch-and-bound solver for\\ One-sided Crossing Minimization}

@@

\newcommand{\mees}[1]{{\color{blue}[{\bf{Bryce:}} #1]}}
\newcommand{\ragnar}[1]{{\color{orange}[{\bf{Ragnar:}} #1]}}

@@

      \begin{alertblock}{todo}
        Figures.
      \end{alertblock}

      \begin{block}{One-sided crossing minimizations}
        Given is a bipartite graph $(A, B)$ that is drawn in the plane at points
        $(i, 0)$ and $(j,1)$ for components $A$ and $B$ respectively. The ordering of $A$
        is fixed. The goal is the find an ordering of $B$ that minimizes the number of
        crossings when edges are drawn as straight lines.

        \heading{Definitions}
        We use $<$ to compare vertices in $A$ in their fixed ordering.

        For $u\in B$, we write $N(u) \subseteq A$ for the set of its neighbours.

        The \textbf{crossing number} $c(u, v)$ is $\sum_{(a, b) \in A^2}
        u(a)v(b)[a > b]$, the number of
        crossings between edges incident to $u$ and $v$ when $u$ is drawn before $v$.  For
        $X,Y\subseteq B$ we set $c(X,Y) = \sum_{x\in X}\sum_{y\in Y} c(x,y)$ for the cost of ordering
        all vertices of $X$ before all vertices of $Y$. More generally, $c(X,Y,Z) = c(X, Y) + c(X, Z) +
        c(Y, Z)$. We also consider the \emph{reduced cost} $r(X,Y) = c(X, Y) - c(Y, X)$, which is
        negative when $X$ is better \emph{before} $Y$ and positive when $X$ is better \emph{after} $Y$.

        We write $u\b v$ when $u$ must come before $v$ in \emph{all} minimal solutions, and say that $(u, v)$
        is a \textbf{fixed pair}.

        A well known result is that when all neighbours of $u$ come before all
        neighbours of $v$, then $u\b v$.

        We try to fix as many pairs as possible to speed up the search.
      \end{block}

      \begin{block}{Strongly fixed pairs}
        We call $u, v \in B$ a \textbf{strongly fixed pair} if their degrees are $n$
        and $m$ and for all $0 \leq i < n$, $N(u)_i \leq N(v)_{\lfloor i\cdot m/n
          \rfloor}$.
        Intuitively, when \emph{the $x\%$'th neighbour of $u$ is left of the $x\%$'th
        neighbour of $v$}. Additionally, at least one of these inequalities must be strict.
        Note that this implies $r(u, v) < 0$.

        \textbf{Main result:}
        If $(u, v)$ is strongly fixed, then $u \b v$.

        \textbf{The lamma is optimal:}
        Suppose that $u, v \in B$ with $N(u) \neq N(v)$ are not strongly fixed, and let $i$ be
        such that $b:=N(u)_i > N(v)_{\lfloor i\cdot m/n\rfloor}=:a$.
        Assuming there exists vertices $m$ in between $a$ and $b$, $l$ left of $\min
        N(v)$ and $r$ right of $\max N(u)$ \ragnar{Is this strong enough? Do $u$ and $v$
          need reversing? Or should we just say left of the min and right of the max
          of the two?}, there exists some \emph{weighted} distribution $X$ over
        $\{l, m, r\}$ such that $c(v, X, u) < \min(c(X, u, v), c(u, v, X))$, so that
        $v$ may have to go before $u$.
      \end{block}

      \begin{block}{Practically fixed pairs}
        Although such a function $X$ may exist in theory, it does
        not have to exist in the actual set $B$, motivating a more practical approach:

        \textbf{Blocking set:}
        Suppose $r(u,v)< 0$, i.e., $u$ wants to go left of $v$.
        A \emph{blocking set} $X\subseteq B-\{u,v\}$ is a set such that $c(v,X,u) \leq \min(c(v, u,
        X), (X, v, u))$.  If there is no blocking set for $(u, v)$, we call it a
        \emph{practically fixed pair}, and $u\b v$.
      \end{block}

@@

      %% Note that we do not require $(v, X, u)$ to be a true local minimum, since we do
      %% not consider interactions between vertices in $X$, as that would make ruling out the
      %% existence of such sets much harder.

@@

    \begin{column}{\colwidth}
      \begin{block}{Gluing}
        A second reduction strategy \emph{glues} vertices $u$ and $v$ together.
        I.e., when $u$ and $v$ are consucutive in \emph{all} optimal solutions,
        we \textbf{glue} them together and treat them as a single vertex.

        \textbf{Gluing identical nodes:}
        When $u$ and $v$ have the same neighbours, we can glue $u$ and $v$.

        \textbf{Greedy gluing:}
        When $r(u, x)\leq 0$ for all $x\in B$, there is a solution that
        starts with $u$.

        \textbf{Practical gluing lemma (unused):}
        Let $u$ and $v$ satisfy $r(u, v) \leq 0$.
        A non-empty subset $X\subseteq B-\{u,v\}$ is \emph{blocking} when $c(u, X, v) \leq
        \min(c(u, v, X), c(X, u, v))$. If there is no blocking set, then we can glue $u, v$.

        Again such sets $X$ can be found or proven to not exist using a knapsack
        algorithm: add points $P_x = (r(u, x), r(x, v))$ and search for a non-empty
        set summing to $\leq{}(0,0)$.
      \end{block}

      \begin{block}{Solver}
        Our solver \texttt{OCMu64} uses a standard \textbf{branch-and-bound on the ordering of the
        solution}.  We start with fixed prefix $P=()$ and tail $T=B$, and in each step we try (a
        subset of) all vertices in $T$ as the next vertex appended to $P$.
        In a preprocessing step we compute the trivial lower bound $S_0 =
        \sum_{u,v}\min(c(u,v),c(v,u))$ on the score.
        %% We keep track of the score $S_P$ of the prefix and $S_{PT}=c(P, T)$ of
        %% prefix-tail intersections, and abort when this score goes above the best
        %% solution found so far.

        \begin{description}
        \item[Graph simplification] We drop degree-$0$ vertices, merge identical
          vertices, and split the graph into \emph{independent} components.
          The initial solution uses the median heuristic and a local search that tries to move
          slices and optimally insert them. We then re-label all nodes in order to make
          memory accesses more efficient.
        \item[Fixed pairs] We find all strongly fixed pairs and store them. For the
          exact track we also find practically fixed pairs.
          For each tail we search for \emph{tail-local} practically fixed pairs. In each state, we only
          try vertices $u\in T$ not fixed by another $v\in T$.
        \item[Gluing] We use the greedy gluing strategy. Our implementation
          of practical gluing contained a bug, so we did not use this.
        \item[Tail cache] In each step, we search for the longest suffix of $T$ that
          was seen before, and reuse (the lower bound on) its score. We also
          cache the tail-local practically fixed pairs.
        \item[Optimal insert] Instead of simply appending $u$ to $P$, we
          insert it in the optimal position. The
          implementation is tricky because it interacts in complicated ways with
          the caching of results for each tail.
        \end{description}
      \end{block}

      \begin{block}{Performance}

      \end{block}
